---
title: "BST260_Project"
author: "Kristyn Beam, Alex Rashedi, and Luis Castelo Branco"
date: "12/13/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(dplyr)
library(lubridate)
library(ggplot2)
library(readr)
library(rvest)
library(ozmaps)
```

## Examining the Effects of Policy Interventions on the Incidence of COVID-19 and Other Infectious Diseases in Australia

### Overview and motivation

COVID-19 has rocked the globe in the year 2020. With it's emergence in Wuhan, China in late 2019, high contagiousness, and lack of available medical treatments and interventions, nonpharmaceutical interventions (NPIs) have been the main defense against its spread. Given this fact, investigation into how NPIs have not only affected the spread of COVID-19, but also other infectious diseases transmitted by physical contact or respiratory droplets, is warranted. 

Influenza is another infectious disease that is tracked annually and globally, and it has a typical course of increased incidence in the winter (colder) months, virtually disappearing in the summer (warmer) months. COVID-19 emerged in December of 2019 but became particularly prevalent in March of 2020. In the United States (US), "influenza season" is typically November to March, where incidence decreases until the following Fall. Given the non-overlapping seasons for high COVID-19 and influenza incidence in the US, in particular, we decided to conduct the investigation using data from a country in the Southern Hemisphere instead, where the periods of high incidence would track more closely. Australia was selected for study, as it has a national data collection system for infectious disease incidence as well as national- and regional-level policy institution for COVID-19 management, lending itself well to analyses of the effects of these policy changes on the incidence of COVID-19 and other infectious diseases of interest. 

### Related work

An [article](https://www-nature-com.ezp-prod1.hul.harvard.edu/articles/s41562-020-01009-0) published recently in Nature examined how specific policy changes (i.e., NPIs) have affected the incidence of COVID-19, globally. 

"The most effective NPIs include curfews, lockdowns and closing and restricting places where people gather in smaller or large numbers for an extended period of time. This includes small gathering cancellations (closures of shops, restaurants, gatherings of 50 persons or fewer, mandatory home working and so on) and closure of educational institutions."

We hypothesize that these NPIs have considerable effects on COVID-19 incidence but may also impact the incidencdes of other infectious diseases.

### Initial questions

Our questions for this project were whether NPIs instituted to limit the spread of COVID-19 also had an effect on other infectious diseases, with a focus on influenza. There are eight states/territories in Australia: New South Wales, Australian Capital Territory, Queensland, Victoria, Tasmania, Southern Australia, Western Australia, and the Northern Territory. We examined several NPIs discussed in the article cited above, including but not limited to: Lockdown, Public Awareness, Health Monitoring, Health Resources, Social Distancing, Restriction of Mass Gatherings, and Closure of Schools. 

Our initial intention was to investigate how the incidence of COVID-19 and influenza changed over time as new policy restrictions were introduced in each region of Australia. As we further explored our data sources, we found that there were several infectious diseases tracked by Australia, allowing us to look at changes in multiple gastrointestinal (GI), respiratory, and sexually transmitted infections (STIs) of interest using both graphical representations and statistical analyses.

Our plan for statistical analysis was to compare the mean incidence rate of both COVID-19 and influenza over the months prior to the specific NPI implemenatation to the mean incidence rate each over the months after NPI implementation. 

### Data sources

We obtained our data from several sources:

1. Australian Infectious Disease Data, located [here](https://www1.health.gov.au/internet/main/publishing.nsf/Content/cdnareport.htm). 

2. Australian COVID-19 data, located [here](need a website here).

3. NPI data, located [here](https://www.coronanet-project.org/). This project's goal is to "...compile a database on government responses to the coronavirus. [The] main focus is to collect as much information as we can about the various fine-grained actions governments are taking to address the effects of the COVID-19 pandemic. To date, this dataset provides the most comprehensive and granular documentation of such government policies in the world, capturing data for 18 broad policy categories alongside many other dimensions, including the initiator, target and timing of a policy."

We downloaded several of the pre-fabricated .csv files from the listed websites. Given that the CoronaNet Research Project is updated frequently, we used webscraping to download this data in order to utilize the most up-to-date information each time our analysis was updated and/or performed. 

We performed numerous data management tasks with the infectious disease data to prepare both count and rate information for analysis, including recoding, filtering, mutating, merging, and transforming. We engaged in similar tasks for the NPI data. Finally, we ensured that all variables across all datasets had the appropriate variable names and type designations for use in our Shiny app. 

### Exploratory data analysis

We first generated exploratory line plots of our infectious disease data to look at counts and rates of COVID-19 and other infectious diseases across time based on province to explore baseline rates for various diseases.

We created a dataframe aus_total_high which restricted the final dataset to include only those diseases with a rate greater than 0.2 (new confirmed cases per million). This removed extremely rare diseases from inclusion in our analysis, and the diseases that were removed with this method were mostly mosquito-borne. 

Lastly, we limited the NPI dataset to only include observations for the first date that each policy type was introduced within each state/territory. Then, we produced test graphs to explore how a vertical line displaying the date for the start of each of these policies visually mapped onto infectious disease incidence within each region.

```{r}
# Read in weekly influenza count data by Australian region 
aus_dat <- read.csv("https://raw.githubusercontent.com/kristynbeam/BST260Project_KB_LC_AR/main/project_data_files/aus_flu_08_17.csv")

# Identify earliest date available: 2017-09-15
min(aus_dat$week)

# Identify latest date available: 2018-01-05
max(aus_dat$week)

aus_dat <- aus_dat %>% 
  # Rename variables
  rename(week = Week.Ending..Friday., 
         age_group = Age..group, 
         indig = Indigenous.status, 
         flu_type = Type.Subtype, 
         state = State, 
         sex = Sex) %>% 
  # Select variables
  select(week, 
         state, 
         age_group, 
         sex, 
         indig, 
         flu_type)

# Convert variable type
aus_dat$week <- mdy(aus_dat$week)

# View aus_dat
head(aus_dat)
```

```{r}
# Create count data frame using aus_dat
aus_dat_case <- aus_dat %>% 
  group_by(week, 
           state, 
           age_group, 
           flu_type, sex) %>% 
  summarize(n_cases = n())
```

```{r}
# Read in Australian infectious disease data
aus_data <- read.csv("../project_data_files/Australian_data_clean.csv")

# Create function to remove NA values
not_all_na <- function(x) any(!is.na(x))

aus_data <- aus_data %>%
  # Remove NA values
  select_if(not_all_na) %>%
  # Rename variable
  rename("Disease" = "X")

# View aus_data
head(aus_data)

# Create case data frame and reshape to long format
aus_data_cases <- aus_data %>% 
  select("Disease":"Aust_n") %>% 
  rename("ACT" = "ACT_n", 
         "NSW" = "NSW_n", 
         "NT" = "NT_n", 
         "Qld" = "Qld_n", 
         "SA" = "SA_n", 
         "Tas" = "Tas_n", 
         "Vic" = "Vic_n", 
         "WA" = "WA_n", 
         "Australia" = "Aust_n") %>% 
  arrange("Date") %>% 
  gather(State, 
         Cases, 
         c("ACT":"Australia"))

# Create rate data frame and reshape to long format
aus_data_rates <- aus_data %>% 
  select("Disease"|"Date"|"ACT_rate":"Aust_rate") %>% 
  rename("ACT" = "ACT_rate", 
         "NSW"="NSW_rate", 
         "NT"="NT_rate", 
         "Qld"="Qld_rate", 
         "SA"="SA_rate", 
         "Tas"= "Tas_rate", 
         "Vic"="Vic_rate", 
         "WA"="WA_rate", 
         "Australia"="Aust_rate") %>% 
  arrange("Date") %>% 
  gather(State, 
         Rates, 
         c("ACT":"Australia")) 

# Join cases and rates data frames
aus_total <- left_join(aus_data_cases, 
                       aus_data_rates)

# Convert variable types
aus_total$Date <- mdy(aus_total$Date)
aus_total$Cases <- as.numeric(aus_total$Cases)
aus_total$Rates <- as.numeric(aus_total$Rates)

# View aus_total
head(aus_total)

# Rename states/territories
aus_total$State <- recode(aus_total$State, 
                          "ACT" = "Australian Capital Territory", 
                          "NSW" = "New South Wales", 
                          "NT" = "Northern Territory", 
                          "Qld" = "Queensland", 
                          "SA" = "South Australia", 
                          "Tas" = "Tasmania", 
                          "Vic" = "Victoria", 
                          "WA" = "Western Australia")

# Filter aus_total data frame to only include diseases that have rate > 0.2
aus_total_high <- aus_total %>% 
  filter(Rates >= 0.2,
         State != "Australia")

# How many diseases are we left with in this case? n = 42
aus_total_high %>% 
  group_by(Disease) %>% 
  summarize()

# Explore using ggplot
aus_total_high %>% 
  ggplot(aes(Date, 
             Rates, 
             color = Disease)) +
  geom_line()

# Save aus_total_high so it can be brought into Shiny app
write.csv(aus_total_high, 
          "aus_total_high.csv")

# View aus_total_high
head(aus_total_high)
```

```{r}
# Create COVID-19 data frame
aus_total_covid <- aus_total %>% 
  filter(Disease == "COVID-19")

# Save aus_total_covid so it can be brought into Shiny app
write.csv(aus_total_covid, 
          "aus_total_covid.csv")

# Create test COVID-19 plots
p1 <- ggplot(data = aus_total_covid, 
            aes(x = Date, 
                y = Cases))

p1 + geom_line(aes(color = State)) +
  ggtitle("COVID-19 Cases by Region") +
  ylim(0, 70000)

p2 <- ggplot(data = aus_total_covid, 
             aes(x = Date, 
                 y = Rates))

p2 + geom_line(aes(color = State)) +
  ggtitle("Covid-19 Rates by Region") +
  ylim(0, 450)

# Create influenza data frame (monthly counts and rates by Australian region)
aus_total_influenza <- aus_total %>% 
  filter(Disease == "Influenza (laboratory confirmed)")

# Identify earliest date available: 2018-11-30
min(aus_total_influenza$Date)

# Identify latest date available: 2020-10-31
max(aus_total_influenza$Date)

# Create test influenza plots
p3 <- ggplot(data = aus_total_influenza, 
             aes(x = Date, 
                 y = Cases))

p3 + geom_line(aes(color = State)) +
  ggtitle("Influenza Cases by Region") +
  ylim(0, 70500)

p4 <- ggplot(data = aus_total_influenza, 
             aes(x = Date, 
                 y = Rates))

p4 + geom_line(aes(color = State)) +
  ggtitle("Influenza Rates by Region") +
  ylim(0, 450)
```

```{r}
# Read in Australian policy data by pulling most up-to-date version from Github
pol_dat_aus = read_csv("https://raw.githubusercontent.com/saudiwin/corona_tscs/master/data/CoronaNet/data_country/coronanet_release/coronanet_release_Australia.csv")

pol_dat_aus <- pol_dat_aus %>%
  # Group pol_dat_aus by province and policy type
  group_by(province, 
           type) %>%
  # Arrange by date policy goes into effect
  arrange(date_start) %>%
  # Keep only first observation within each province-policy type group
  slice(1) %>%
  # Ungroup
  ungroup() %>%
  # Filter to keep only policies that target Australia
  filter(target_country == "Australia")

# Transform variable types
pol_dat_aus$type <- as.factor(pol_dat_aus$type)
pol_dat_aus$type_sub_cat <- as.factor(pol_dat_aus$type_sub_cat)
pol_dat_aus$target_who_what <- as.factor(pol_dat_aus$target_who_what)

# View pol_dat_aus
head(pol_dat_aus)
list.files()

# Save pol_dat_aus so it can be brought into Shiny app
write.csv(pol_dat_aus, "pol_dat_aus.csv")
```

```{r}
# https://www-nature-com.ezp-prod1.hul.harvard.edu/articles/s41562-020-01009-0
# "The most effective NPIs include curfews, lockdowns and closing and restricting places where people gather in smaller or large numbers for an extended period of time. This includes small gathering cancellations (closures of shops, restaurants, gatherings of 50 persons or fewer, mandatory home working and so on) and closure of educational institutions."

# Test graph
aus_total_high %>%
  filter(State == "New South Wales" & Date >= "2020-01-01" & Disease %in% c("Influenza (laboratory confirmed)",
                                                                            "COVID-19")) %>%
  ggplot(aes(x = Date, 
             y = Rates)) +
  geom_line(aes(color = Disease), 
            alpha = 0.5) +
  # 2020-03-16 is date of first policy restriction in NSW (restriction of mass gatherings)
  geom_vline(alpha = 0.5, 
             xintercept = as.numeric(as.Date("2020-03-16")), 
             color = "blue") +
  labs(x = "Date",
       y = "Rate") +
  ggtitle("Covid rates in NSW over time") +
  scale_x_date(date_breaks = "1 month", 
               date_labels = "%b")
```

### Statistical analysis

```{r}
# Transform variable to date type
aus_total_influenza$Date <- ymd(aus_total_influenza$Date)

# View aus_total_influenza
head(aus_total_influenza)

# Create a variable: drop in influenza cases from year 1 (Nov/18 to Oct/19) to year 2 (Nov19 to Oct/20)
aus_total_influenza_year1 <- aus_total_influenza %>% 
  filter(Date <= "2019-10-31") %>%
  group_by(State) %>%
  summarize(year1 = sum(Cases))

aus_total_influenza_year2 <- aus_total_influenza %>% 
  filter(Date > "2019-10-31") %>%
  group_by(State) %>%
  summarize(year2 = sum(Cases))

aus_total_influenza_diff <- left_join(aus_total_influenza_year1,
                                      aus_total_influenza_year2) %>%
  mutate(drop = (year1 - year2) / year1)
```


```{r}
# Perform web scraping for a dataset that containd the population and population density for the provinces
url <- "https://www.worldatlas.com/articles/which-australian-states-territories-have-the-highest-density-of-population.html"
h <- read_html(url)
density <- h %>% 
  html_nodes("table") %>%
  .[1]
density <- density %>% 
  html_table %>% 
  .[[1]] 

density <- density %>%
  rename("State" = "State / Territory") %>%
  rename("density" = "Density (persons per km2)") %>% 
  mutate(State = recode(State, 
                        "Australian Capital Territory" = "ACT", 
                        "New South Wales" = "NSW", 
                        "Northern Territory" = "NT", 
                        "Queensland" = "Qld", 
                        "South Australia" = "SA", 
                        "Tasmania" = "Tas", 
                        "Victoria" = "Vic", 
                        "Western Australia" = "WA"))

density$Population <- as.numeric(gsub(",", 
                                      "", 
                                      density$Population))

# Join the datasets
aus_total_influenza_diff <- left_join(aus_total_influenza_diff, 
                                      density)

# View aus_total_influenza_diff
head(aus_total_influenza_diff)
```

```{r}
# Create map of Australia
x <- ozmap_data()
ggplot(data = x) + 
  geom_sf()

# This is an "sf" data structure which can be treated as a df but with an additional column called a geometry. If we want to color the data by Cases or Rates, we would merge this with our df that has the cases and rates information.

# Create a merged dataset of x and aus_total_high
merge_dat <- x %>% 
  rename(State = NAME) %>% 
  merge(aus_total_high, 
        by = "State")
```

#### Linear regressions and t-tests

```{r}
# Build statistical tests to see if the mean cases of a disease in the first year (without COVID-19) was different from the mean cases in the second year (with COVID-19)

# Example with Influenza
aus_total_influenza_year1 <- aus_total_high %>% 
  filter(Disease == "Influenza (laboratory confirmed)") %>%
  filter(Date <= "2019-10-31") %>%
  group_by(State) %>%
  summarize(year1 = mean(Cases))

aus_total_influenza_year2 <- aus_total_high %>%
  filter(Disease == "Influenza (laboratory confirmed)") %>%
  filter(Date > "2019-10-31") %>%
  group_by(State) %>%
  summarize(year2 = mean(Cases))

aus_total_influenza_test <- left_join(aus_total_influenza_year1, 
                                      aus_total_influenza_year2) %>%
  filter(State != "Australia")

t.test(aus_total_influenza_test$year1, 
       aus_total_influenza_test$year2, 
       paired = TRUE, 
       alternative = "two.sided")

# Example with Gonococcal infection
aus_total_Gono_year1 <- aus_total_high %>%
  filter(Disease == "Gonococcal infection") %>%
  filter(Date <= "2019-10-31") %>%
  group_by(State) %>%
  summarize(year1 = mean(Cases))

aus_total_Gono_year2 <- aus_total_high %>%
  filter(Disease == "Gonococcal infection") %>%
  filter(Date > "2019-10-31") %>%
  group_by(State) %>%
  summarize(year2 = mean(Cases))

aus_total_Gono_test <- left_join(aus_total_Gono_year1, 
                                 aus_total_Gono_year2) %>%
  filter(State != "Australia")

t.test(aus_total_Gono_test$year1, 
       aus_total_Gono_test$year2, 
       paired = TRUE, 
       alternative = "two.sided")

# Example with Hepatitis C
aus_total_HepC_year1 <- aus_total_high %>%
  filter(Disease == "Hepatitis C (unspecified)") %>%
  filter(Date <= "2019-10-31") %>%
  group_by(State) %>%
  summarize(year1 = mean(Cases))

aus_total_HepC_year2 <- aus_total_high %>%
  filter(Disease == "Hepatitis C (unspecified)") %>%
  filter(Date > "2019-10-31") %>%
  group_by(State) %>%
  summarize(year2 = mean(Cases))

aus_total_HepC_test <- left_join(aus_total_HepC_year1, 
                                 aus_total_HepC_year2) %>%
  filter(State != "Australia")

t.test(aus_total_HepC_test$year1, 
       aus_total_HepC_test$year2, 
       paired = TRUE, 
       alternative = "two.sided")
  
# Example with Rotavirus
aus_total_Rota_year1 <- aus_total_high %>% 
  filter(Disease == "Rotavirus") %>%
  filter(Date <= "2019-10-31") %>%
  group_by(State) %>%
  summarize(year1 = mean(Cases))

aus_total_Rota_year2 <- aus_total_high %>% 
  filter(Disease == "Rotavirus") %>%
  filter(Date > "2019-10-31") %>%
  group_by(State) %>%
  summarize(year2 = mean(Cases))

aus_total_Rota_test <- left_join(aus_total_Rota_year1, 
                                 aus_total_Rota_year2) %>%
  filter(State != "Australia")

t.test(aus_total_Rota_test$year1, 
       aus_total_Rota_test$year2, 
       paired = TRUE, 
       alternative = "two.sided")

# Example with Pertussis
aus_total_Pertussis_year1 <- aus_total_high %>% 
  filter(Disease == "Pertussis") %>%
  filter(Date <= "2019-10-31") %>%
  group_by(State) %>%
  summarize(year1 = mean(Cases))

aus_total_Pertussis_year2 <- aus_total_high %>% 
  filter(Disease == "Pertussis") %>%
  filter(Date > "2019-10-31") %>%
  group_by(State) %>%
  summarize(year2 = mean(Cases))

aus_total_Pertussis_test <- left_join(aus_total_Pertussis_year1, 
                                      aus_total_Pertussis_year2) %>%
  filter(State != "Australia")

t.test(aus_total_Pertussis_test$year1, 
       aus_total_Pertussis_test$year2, 
       paired = TRUE, 
       alternative = "two.sided")
```

### Final conclusions
