---
title: "BST260 Final Project"
author: "Kristyn Beam, Luis Castelo Branco, Alex Rashedi"
date: "12/13/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE}
#Loading packages necessary for the analysis
library(tidyverse)
library(dplyr)
library(lubridate)
library(ggplot2)
library(readr)
```

## Examining Policy Interventions on COVID-19 and Other Infectious Disease Incidence in Australia

### Overview and Motivation

COVID-19 has rocked the globe in the year 2020. With it's emergence in late 2019, quick spread, and lack of medical treatments and interventions, nonpharmaceutical interventions have been the forefront of limiting spread, disease, and death. 



### Related work

An [article](https://www-nature-com.ezp-prod1.hul.harvard.edu/articles/s41562-020-01009-0) published recent in Nature examined how policy changes, specifically nonpharamaceutical interventions (NPIs) affected the incidence of COVID globally. 

"The most effective NPIs include curfews, lockdowns and closing and restricting places where people gather in smaller or large numbers for an extended period of time. This includes small gathering cancellations (closures of shops, restaurants, gatherings of 50 persons or fewer, mandatory home working and so on) and closure of educational institutions."



### Initial questions




### Data sources





### Exploratory Analysis




### Final Analysis

