---
title: "BST260 Final Project"
author: "Kristyn Beam, Luis Castelo Branco, Alex Rashedi"
date: "12/13/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE}
#Loading packages necessary for the analysis
library(tidyverse)
library(dplyr)
library(lubridate)
library(ggplot2)
library(readr)
```

## Examining Policy Interventions on COVID-19 and Other Infectious Disease Incidence in Australia

### Overview and Motivation

COVID-19 has rocked the globe in the year 2020. With it's emergence in late 2019, quick spread, and lack of medical treatments and interventions, nonpharmaceutical interventions have been the forefront of limiting spread, disease, and death. Given the fact that non-pharmaceutical interventions (NPIs) have been the main defense against spread of COVID-19, investigation into how NPIs have not only affected the spread of COVID-19, a respiratory disease, but also other infectious diseases spread by contact or respiratory droplets. 

The other main infectious disease tracked annually and globally is influenza which has a typical course of increased incidence in the winter (colder) months and virtually disappears in the summer (warmer) months. COVID emerged in December of 2019, but became particularly prevalent in March of 2020. In the United States, the season for influenza is typically November to March and then influenza typically decreases in incidence until the following Fall. Given the alternating seasons for tracking COVID and influenza in particular in the United States, we wanted to see the effects of NPIs on both COVID incidence and also other respiratory infections. To do this, we needed to investigate rates in a country where the seasons of COVID and influenza would be tracking along similarly, which was the Southern Hemisphere. Australia has a national data collection system for both infectious diseases as well as national policy institution for COVID management lending itself well to analysis of these policies on multiple infectious diseases. 

### Related work

An [article](https://www-nature-com.ezp-prod1.hul.harvard.edu/articles/s41562-020-01009-0) published recently in Nature examined how policy changes, specifically nonpharamaceutical interventions (NPIs) affected the incidence of COVID globally. 

"The most effective NPIs include curfews, lockdowns and closing and restricting places where people gather in smaller or large numbers for an extended period of time. This includes small gathering cancellations (closures of shops, restaurants, gatherings of 50 persons or fewer, mandatory home working and so on) and closure of educational institutions."

It appears that these NPIs have variable effects on COVID-19 incidence but may also have affected other infectious disease incidence.


### Initial questions

Our questions for this project were whether non-pharmaceutical interventions meant to limit the spread of COVID-19 also had an effect on other infectious diseases, with a focus on influenza. 

There are 8 provinces in Australia: New South Wales, Australian Capital Territory, Queensland, Victoria, Tasmania, Southern Australia, Western Australia, and the Northern Territory. We examined several NPIs studied through the article cited above including: Lockdown, Public Awareness, Health Monitoring, Health Resources, Social Distancing, Restriction of Mass Gatherings, and Closure of Schools as a few examples. 

We explored this question of changing incidence first visually through graphical representations of the disease cases and rates with a vertical line indicating when certin policies were implemented in each province in Australia.

We examined if there was a statistically significant difference between the incidence of certain infectious diseases (influenza) pre NPI implementation and post NPI implementation.


### Data sources

We obtained our data from several data sources for completeness.

1. Australian Infectious Disease Data, located [here](https://www1.health.gov.au/internet/main/publishing.nsf/Content/cdnareport.htm). 
2. Australian COVID data, located [here](need a website here).
3. Non-pharmaceutical policy interventions data. This data came from the [CoronaNet Research Project](https://www.coronanet-project.org/). This project's goal is to: "...compile a database on government responses to the coronavirus. Our main focus is to collect as much information as we can about the various fine-grained actions governments are taking to address the effects of the COVID-19 pandemic. To date, this dataset provides the most comprehensive and granular documentation of such government policies in the world, capturing data for 18 broad policy categories alongside many other dimensions, including the initiator, target and timing of a policy."


### Exploratory Data Analysis

```{r}
# Read in data on weekly influenza counts by Australian region 
aus_dat <- read.csv("../project_data_files/aus_flu_08_17.csv")

# Identify earliest date available: 2017-09-15
min(aus_dat$week)

# Identify latest date available: 2018-01-05
max(aus_dat$week)

aus_dat <- aus_dat %>% 
  # Rename variables
  rename(week = Week.Ending..Friday., age_group = Age..group, indig = Indigenous.status, flu_type = Type.Subtype, state = State, sex = Sex) %>% 
  # Select variables
  select(week, state, age_group, sex, indig, flu_type)

# Convert variable types
aus_dat$week <- mdy(aus_dat$week)
# aus_dat$flu_type <- as.character(aus_dat$flu_type)
# aus_dat$state <- as.character(aus_dat$state)
# aus_dat$sex <- as.character(aus_dat$sex)

# View aus_dat
head(aus_dat)

# Create count data frame using aus_dat
aus_dat_case <- aus_dat %>% 
  group_by(week, state, age_group, flu_type, sex) %>% 
  summarize(n_cases = n())
```


```{r}
# Read in Australian disease data
aus_data <- read.csv("../project_data_files/Australian_data_clean.csv")

# removing all NA values
not_all_na <- function(x) any(!is.na(x))

aus_data <- aus_data %>%
  select_if(not_all_na) %>%
  # Rename variable
  rename("Disease"="X")

head(aus_data)

# Create case data frame and reshape to long format
aus_data_cases <- aus_data %>% 
  select("Disease":"Aust_n") %>% 
  rename("ACT"="ACT_n", "NSW"="NSW_n", "NT"="NT_n", "Qld"="Qld_n", "SA"="SA_n", "Tas"= "Tas_n", "Vic"= "Vic_n", "WA"= "WA_n", "Australia"= "Aust_n") %>% 
  arrange("Date") %>% 
  gather(State, Cases, c("ACT":"Australia"))

# Create rate data frame and reshape to long format
aus_data_rates <- aus_data %>% 
  select("Disease"|"Date"|"ACT_rate":"Aust_rate") %>% 
  rename("ACT"="ACT_rate", "NSW"="NSW_rate", "NT"="NT_rate", "Qld"="Qld_rate", "SA"="SA_rate", "Tas"= "Tas_rate", "Vic"="Vic_rate", "WA"="WA_rate", "Australia"="Aust_rate") %>% 
  arrange("Date") %>% 
  gather(State, Rates, c("ACT":"Australia")) 

# Join cases and rates data frames
aus_total <- left_join(aus_data_cases, aus_data_rates)

# Convert variable types
aus_total$Date <- mdy(aus_total$Date)
aus_total$Cases <- as.numeric(aus_total$Cases)
aus_total$Rates <- as.numeric(aus_total$Rates)

# View aus_total
head(aus_total)

# Rename provinces
aus_total$State <- recode(aus_total$State, "ACT" = "Australian Capital Territory", "NSW" = "New South Wales", "NT" = "Northern Territory", "Qld" = "Queensland", "SA" = "South Australia", "Tas" = "Tasmania", "Vic" = "Victoria", "WA" = "Western Australia")


#Filter aus_total df to only include diseases that have a rate of disease of > 0.2/100,000 
aus_total_high <- aus_total %>% 
  filter(Rates >= 0.2)

#How many diseases are we left with in this case? n = 42
aus_total_high %>% 
  group_by(Disease) %>% 
  summarize()

head(aus_total_high)
write.csv(aus_total_high, "../project_data_files/aus_total_high.csv")
```

```{r}
# Create COVID-19 data frame
aus_total_covid <- aus_total %>% 
  filter(Disease == "COVID-19")

write.csv(aus_total_covid, "../project_data_files/aus_total_covid.csv")

# Create test COVID-19 plots
p1 <- ggplot(data = aus_total_covid, 
            aes(x = Date, 
                y = Cases))
p1 + geom_line(aes(color = State)) +
  ggtitle("Covid Cases by Region") +
  ylim(0, 70000)

p2 <- ggplot(data = aus_total_covid, 
             aes(x = Date, 
                 y = Rates))
p2 + geom_line(aes(color = State)) +
  ggtitle("Covid Rates by Region") +
  ylim(0, 450)

# Create influenza data frame (monthly counts and rates by Australian region)
aus_total_influenza <- aus_total %>% 
  filter(Disease == "Influenza (laboratory confirmed)")

# Identify earliest date available: 2018-11-30
min(aus_total_influenza$Date)

# Identify latest date available: 2020-10-31
max(aus_total_influenza$Date)

# Create test influenza plots
p3 <- ggplot(data = aus_total_influenza, 
             aes(x = Date, 
                 y = Cases))
p3 + geom_line(aes(color = State)) +
  ggtitle("Influenza Cases by Region") +
  ylim(0, 70500)

p4 <- ggplot(data = aus_total_influenza, 
             aes(x = Date, 
                 y = Rates))
p4 + geom_line(aes(color = State)) +
  ggtitle("Influenza Rate by Region") +
  ylim(0, 450)
```


```{r}
# Read in Australian policy data by pulling most up-to-date version from Github
pol_dat_aus = read_csv("https://raw.githubusercontent.com/saudiwin/corona_tscs/master/data/CoronaNet/data_country/coronanet_release/coronanet_release_Australia.csv")

pol_dat_aus <- pol_dat_aus %>%
  # Group pol_dat_aus by province and policy type
  group_by(province, 
           type) %>%
  # Arrange by date policy goes into effect
  arrange(date_start) %>%
  # Keep only first observation within each province-policy type group
  slice() %>%
  # Ungroup
  ungroup() %>%
  # Filter to keep only policies that target Australia
  filter(target_country == "Australia")

pol_dat_aus$type <- as.factor(pol_dat_aus$type)
pol_dat_aus$type_sub_cat <- as.factor(pol_dat_aus$type_sub_cat)
pol_dat_aus$target_who_what <- as.factor(pol_dat_aus$target_who_what)

# View pol_dat_aus
head(pol_dat_aus)

# Save pol_dat_aus so it can be brought into Shiny app
write.csv(pol_dat_aus, "../project_data_files/pol_dat_aus.csv")

#Recode province names to match other dataframes
country_names <- recode(pol_dat_aus$province, "ACT" = "Australian Capital Territory", "NSW" = "New South Wales", "NT" = "Northern Territory", "Qld" = "Queensland", "SA" = "South Australia",  "Tas" = "Tasmania", "Vic" = "Victoria", "WA" = "Western Australia")

pol_dat_aus$province <- country_names
```


```{r}
# Test graph for shiny app
aus_total_high %>%
  filter(State == "New South Wales" & #Date >= "2020-01-01" & 
           Disease %in% c("Influenza (laboratory confirmed)", "COVID-19")) %>%
  ggplot(aes(x = Date, 
             y = Rates)) +
  geom_line(aes(color = Disease), 
            alpha = 0.5) +
  # 2020-03-16 is date of first policy restriction in NSW (restriction of mass gatherings)
  geom_vline(alpha = 0.5, 
             xintercept = as.numeric(as.Date("2020-03-16")), 
             color = "blue") +
  labs(x = "Date",
       y = "Rate") +
  ggtitle("Covid rates in NSW over time") +
  scale_x_date(date_breaks = "1 month", 
               date_labels = "%b")
```


### Statistical Analysis

```{r}
#Creating a variable: drop in influenza cases from year 1 (Nov/18 to Oct/19) compared to year 2 (Nov19 to Oct/20)
aus_total_influenza$Date <- ymd(aus_total_influenza$Date)

head(aus_total_influenza)

aus_total_influenza_year1<- aus_total_influenza %>% filter (Date <="2019-10-31") %>%group_by(State)%>%summarize(year1=sum(Cases))
aus_total_influenza_year2<- aus_total_influenza %>% filter (Date > "2019-10-31") %>%group_by(State)%>%summarize(year2=sum(Cases))
aus_total_influenza_diff<- left_join(aus_total_influenza_year1, aus_total_influenza_year2)%>%mutate(drop=(year1-year2)/year1)

```


```{r}
#web scraping for a dataset that containd the Population and Population density for the provinces
library(rvest)
url <- "https://www.worldatlas.com/articles/which-australian-states-territories-have-the-highest-density-of-population.html"
h <- read_html(url)
density<- h %>% html_nodes("table") %>% .[1]
density <- density %>% html_table %>% .[[1]] 

density<- density%>%rename("State"= "State / Territory")%>%rename("density"="Density (persons per km2)") %>% mutate(State=recode(State,"Australian Capital Territory"="ACT", "New South Wales"="NSW", "Northern Territory"="NT", "Queensland"="Qld", "South Australia"="SA", "Tasmania"="Tas", "Victoria"="Vic", "Western Australia"="WA"))

density$Population<-as.numeric(gsub(",","",density$Population))

#joining the datasets
aus_total_influenza_diff<-left_join(aus_total_influenza_diff, density)

head(aus_total_influenza_diff)

#linear regressions to see if population or population density were associated with the drop in cases

# linearMod <- lm(drop ~ density , data=aus_total_influenza_diff) 
# confint.lm(linearMod)
# summary(linearMod)
# 
# linearMod2 <- lm(drop ~ Population , data=aus_total_influenza_diff) 
# confint.lm(linearMod2)
# summary(linearMod2)
```

### Final Conclusions


